 

### 2021-05-20 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC 2021), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [*Trenberth et al. 2014*].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities]. Leading to, individually or in combination, reduced forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].

@@

Stomatal conductance ($g_s$) is a key factor mediating $T_{leaf}$ [@songTreeSurfaceTemperature2020] because leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because $gs$ facilitates heat dissipation, when water demand exceeds water transport, driven by for example: irradiance,$T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.

@@

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
 Inherent leaf $gs$ is an important factor for $T_{leaf}$ which is typically greater across species exposed sun-leaves due to stomata density than shade leaves. However, because factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress.

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves tend to heat and cool faster due to higher $ga$, maintaining lower $T_{leaf}$-$T_{air}$ than large leaves that take longer to heat and cool due to low $ga$ [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017].

Lastly, $gs$ and leaf size together are most important factors in deciding $T_{leaf}$ - $T_{air}$ and temperature sensitivity, because $gs$ determines leaf biological-biophysical interaction and leaf size determines leaf heating and cooling rate. Therefore, smaller leaves with greater $gs$ have are less sensitive to high temperature [@songTreeSurfaceTemperature2020].

@@

@@

Across forest species, widely, show fundamental trait adaptation to light gradients similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds due to inherent specie-specific traits. Spatio-temporal gradients offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007]. Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. Species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase [@fauset_differences_2018; @songTreeSurfaceTemperature2020].

@@

Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [Oishi et al. 2016].This can implication to canopies and gaps that reside at levels of greater $VPD$. Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. However, soil water loss during drought effects larger trees more than small trees because, limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, bennett; @mcgregor_tree_nodate].

@@

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).
It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. Interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height, light and water. Partly because NSC reserves are greater in taller trees due to available light than in buffered trees that are light and NSC limited, therefore making buffered small trees more sensitive to stress [@niinemetsResponsesForestTrees2010a].

@@

While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]. Warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to fauna in microrefugias [@scheffersMicrohabitatsReduceAnimal2013] . Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020]. Whereas, in other cases warming is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020]. Heatwave and temperature  can alter canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019]. At a leaf metabolism level, fluxnet global data analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021].

>>>>

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC 2021), accompanied by increasing severity and frequency of heat waves [@perkinsIncreasingFrequencyIntensity2012] and hotter droughts [*Trenberth et al. 2014*].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function (@breshearsUnderappreciatedPlantVulnerabilities; IPBES report), including altered rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities 2021; @corlettImpactsWarmingTropical2011], foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reduced growth and elevated mortality during drought [@teskeyResponsesTreeSpecies2015; @breshearsUnderappreciatedPlantVulnerabilities]. Leading to, individually or in combination, reduced forest C sequestration, as may already be occurring in some forests [REFS; @sullivanLongtermThermalSensitivity2020].The resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future trajectory of climate change [@bonan_ecological_2016, Bonan 2008], yet this influence remains extremely uncertain [REFS].

@@

Stomatal conductance ($g_s$) is a key factor mediating $T_{leaf}$ [@songTreeSurfaceTemperature2020].Leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because $gs$ facilitates heat dissipation, when water demand exceeds water transport, driven by for example: irradiance,$T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.

@@

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
 Inherent leaf $gs$ is typically greater across species exposed sun-leaves due to stomata density than shade leaves. However, because factors driving $g_s$ limitation--irradiance, VPD and $T_{air}$--increase with exposure, $g_s$ limitation is steeper for sun- than shade-leaves, driving an increase in $T_{leaf}$ (@leighInfluenceLeafSize2017; @fauset_differences_2018). On the other hand, shade leaves can remain cooler under buffered conditions, but lower leaf $gs$ can result in greater $T_{leaf}$ under slight heat stress.

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves tend to heat and cool faster due to higher $ga$, maintaining lower $T_{leaf}$-$T_{air}$ than large leaves that take longer to heat and cool due to low $ga$, therefore can experience greater $T_{leaf}$ - $T_{air}$ than small leaves [@bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020].

Lastly, $gs$ and leaf size together, though donot solely explain mean $T_{leaf}$, are most important factors in deciding $T_{leaf}$ - $T_{air}$ and temperature sensitivity, because $gs$ determines leaf biological-biophysical interaction and leaf size determines leaf heating and cooling rate. Therefore, smaller leaves with greater $gs$ are less sensitive to high temperature [@songTreeSurfaceTemperature2020].

@@

-"Leaves with adequate water supply are partly buffered against air-temperature
variations, through a passive biophysical mechanism"[@dongBiophysicalHomoeostasisLeaf2017].

@@

Across forest species, widely, show fundamental trait adaptation to light gradients similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds due to inherent specie-specific traits. Spatio-temporal gradients offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007]. Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000]. Therefore $T_{leaf}$ varies across speci-specific traits, where species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase [@fauset_differences_2018; @songTreeSurfaceTemperature2020].

@@

Biophysically, greater $VPD$, atmospheric demand and low RH can be more stressful for trees, than soil moisture across biomes-including mesic forests,  as the former limits surface conductance and evapotranspiration to a greater extent than soil moisture [Oishi et al. 2016; @corlettImpactsWarmingTropical2011] .This can implication to canopies and gaps that reside at levels of greater $VPD$. Whereas, greater $RH$ can do the opposite, where it stifles evapotranspiration thereby increasing $T_{leaf}$, particularly in tropical canopies where irradiance can have a greater effect [Fig.3; @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979]. However, soil water loss during drought effects larger trees more than small trees because, limited soil-water cannot meet the evaporative demand via the long-distance water transport to the canopy [Norway spruce study, bennett; @mcgregor_tree_nodate].

@@

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).
It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. 
-Whereas, in other cases warming is associated with taller understory trees with lower LMA that complete their lifecycle in the understory, under closed-canopy temperate forests [@maesPlantFunctionalTrait2020]
-Decelarating tree growth rates predominantly in smaller trees, but not restricted to, have also been observed in tropical forests, associated with local climate patterns of mean daily minimum temperature, increasing temperature and rainfall regime [Feeley et al. 2007] 
-Interacting biophysical and biological stressors also show that stress sensitivity and ability to tolerate number of stressors increases with height, light and water. Partly because NSC reserves are greater in taller trees due to available light than in buffered trees that are light and NSC limited, therefore making buffered small trees more sensitive to stress [@niinemetsResponsesForestTrees2010a].

@@

While forest microclimates are insulated from macroclimatic extremes, they are still subjected to warming trends [@bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]. Warming will disproportionately affect the less warm-adapted functional types within the forest canopy, essentially creating holes in the canopy that will take a long time to refill (especially in slow-growing tropical sub-canopy specialists), as is observed with understory plant thermophilization (death is faster than recruitment and growth) [@duqueThermophilizationAdultJuvenile2015a; @zellweger_forest_2020; @zellweger_seasonal_2019; @defrenneMicroclimateModeratesPlant2013] with increasing threat to fauna in microrefugias [@scheffersMicrohabitatsReduceAnimal2013] . Tree-ring analysis in mesic temperate forests shows reduced growth of understory trees relative to overstory with warming temperatures [@rollinsonClimateSensitivityUnderstory, 2020]. Heatwave and temperature  can alter canopy and understory phenology, forest photosynthesis and forest C uptake [REFS, @zohnerOngoingSeasonallyUneven2019]. At a leaf metabolism level, fluxnet global data analysis shows that with temperature increase, foliage photosynthesis and respiration are coupled at different rates upto photosynthetic thermal optimum, beyond this inflection point, photosynthesis declines whereas respiration increases exponentially, therfore, moving from carbon sink to carbon source, independent of water and irradiance [@duffyHowCloseAre2021].

Tropical forests are narrower thermal limits [@corlettImpactsWarmingTropical2011]
